### 2020-06-24 submission/manuscript.tex

The microbiota plays a key role in determining susceptibility to
\emph{Clostridioides difficile} infections (CDIs). However, much of the
mechanistic work examining CDIs in mouse models use a single university
colony or vendor, which have lower inter-individual microbiota variation
compared to humans. We treated mice from 6 different colony sources (2
University of Michigan colonies and 4 vendors) with a single clindamycin
dose, followed by \emph{C. difficile} challenge 1 day later and measured
\emph{C. difficile} colonization levels through 9 days post-infection.
The microbiota was profiled throughout the experiment via 16S rRNA gene
sequencing analysis to examine variation across colony sources and
alterations due to clindamycin treatment and \emph{C. difficile}
challenge. While all sources of mice were colonized 1-day
post-infection, variation in \emph{C. difficile} colonization levels
emerged from days 3-7 post-infection with 3 sources colonized with
\emph{C. difficile} for longer and at higher levels. We identified
bacterial taxa with different relative abundances across colony sources
throughout the experiment, as well as taxa that were consistently
impacted by clindamycin treatment. We created bacterial community-based
logistic regression models that successfully classified mice based on
their day 7 \emph{C. difficile} colonization status. Importantly, after
examining the taxa that were most important to the classification
models, we identified a subset of key taxa that varied across colony
sources (\emph{Bacteroides}, \emph{Deferribacteraceae}), were altered by
clindamycin (\emph{Porphyromonadaceae}, \emph{Ruminococcaceae}), or both
(\emph{Enterobacteriaceae}, \emph{Enterococcus},
\emph{Bifidobacteriaceae}, \emph{Coriobacteriaceae},
\emph{Lachnospiraceae}, and \emph{Verrucomicrobiaceae}). These results
suggest the response of the initial gut microbiota to clindamycin
treatment influences \emph{C. difficile} 630 colonization dynamics.

@@

Mouse models of CDIs have been a great tool for understanding \emph{C.
difficile} pathogenesis (13). The number of CDI mouse model studies has
grown substantially since Chen et al. published their C57BL/6 model in
2008, which disrupted the gut microbiota with antibiotics to enable
\emph{C. difficile} colonization, diarrhea, and weight loss (14). CDI
mouse models have been used to examine translationally relevant
questions regarding \emph{C. difficile}, including the role of the
microbiota and efficacy of potential therapeutics for treating CDIs
(15). However, microbiome variation between lab mice is much less than
the variation observed between humans (16, 17). Additionally, studying
the contribution of the microbiota to a particular disease phenotype in
one set of lab mice after the same perturbation could yield a number of
findings of which only a fraction may be driving the phenotype.

@@

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization regardless of colony source.} To test how
the microbiotas of mice from different colony sources impact
colonization dynamics after clindamycin exposure, we utilized C57BL/6
mice from 6 different sources: two colonies from the University of
Michigan (the Young and Schloss lab colonies), the Jackson Laboratory,
Charles River Laboratories, Taconic Biosciences, and Envigo (which was
formerly Harlan). These 4 vendors were chosen because they represent
commonly used vendors for CDI studies in mice (26, 34--40). After a
13-day acclimation period for the mice ordered from vendors, all mice
were treated with 10 mg/kg clindamycin via intraperitoneal injection and
one day later challenged with 10\textsuperscript{3} \emph{C. difficile}
630 spores (Fig. 1A). Clindamycin was chosen because we have previously
demonstrated mice are rendered susceptible, but consistently cleared the
CDI within 9 days (21, 41), clindamycin is frequently implicated with
human CDIs (42), and is also part of the antibiotic treatment for the
frequently cited 2008 CDI mouse model (14). The day after infection,
\emph{C. difficile} was detectable in all mice at a similar level (Mean
CFU range: 3e+07-1.5e+08; \emph{P}\textsubscript{FDR} = 0.15),
indicating clindamycin rendered all mice susceptible regardless of
colony source (Fig. 1B). Interestingly, variation in \emph{C. difficile}
CFU levels across sources of mice emerged from days 3-7 post-infection
(all \emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 1B and Table S1),
suggesting mouse colony source is associated with \emph{C. difficile}
clearance. We conducted two experiments approximately 3 months apart and
combined the data because the colonization dynamics across sources of
mice varied similarily in both experiments (Fig. S1). Although \emph{C.
difficile} 630 causes mild symptoms in mice comparied to other \emph{C.
difficile} strains (43), we also saw that weight change significantly
varied across sources of mice with the most weight lost two days
post-infection (Fig. 1C and Table S2). Interestingly, mice ordered from
Jackson, Taconic, and Envigo tended to lose the most weight, have higher
\emph{C. difficile} CFU levels and take longer to clear the infection
compared to the other sources of mice, which was particularly evident 7
days post-infection (Fig. 1C and D). Importantly, there was also one
Jackson and one Envigo mouse that died between 1- and 3-days
post-infection. Thus, clindamycin rendered all mice susceptible to
\emph{C. difficile} 630 colonization, regardless of colony source, but
variation across sources emerged with 3 out of 6 sources taking longer
to clear \emph{C. difficile}.

\textbf{Bacterial communities vary across mouse colony sources.} Given
the well known variation in mouse microbiomes across vendors and
university colonies (25), we hypothesized that the variation in \emph{C.
difficile} clearance could be explained by microbiota variation across
the 6 sources. We used 16S rRNA gene sequencing to characterize the
fecal bacterial communities from the mice over the course of the
experiment. Since antibiotics and other risk factors of CDIs are
associated with decreased microbiota diversity (44), we first examined
alpha diversity measures across the 6 sources of mice. Examining the
bacterial communities at baseline, prior to clindamycin treatment there
was a significant difference in the number of observed OTUs
(\emph{P}\textsubscript{FDR} = 0.03), but not Shannon diversity index
(\emph{P}\textsubscript{FDR} = 0.052) across sources of mice (Fig. 2A-B
and Table S3). As expected, clindamycin treatment decreased richness and
Shannon diversity across all sources of mice, and communities started to
recover 1 day post-infection (Fig. 2C-D). Interestingly, significant
differences in diversity metrics across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after both
clindamycin and \emph{C. difficile} infection, with Charles River mice
having higher richness and Shannon Diversity than most of the other
groups (Fig 2C-F and Table S4). While Charles River mice had more
diverse microbiotas, Young and Schloss lab mice were also able to clear
\emph{C. difficile} faster, suggesting microbiota diversity alone does
not explain the observed variation in \emph{C. difficile} colonization
across vendors.

Next, we compared the bacterial communities from the 6 colonies over the
course of the experiment using principal coordinate analysis (PCoA) of
the Theta YC distances. Permutational multivariate analysis of variance
(PERMANOVA) analysis revealed colony source was the major factor
explaining the observed variation across fecal communities
(R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
interactions between cage and day of the experiment (Movie S1 and Table
S5). Since, the majority of the perturbations happened over the initial
days of the experiment, we decided to focus on the bacterial communities
at baseline (day -1), after clindamycin treatment (day 0), and
post-infection (day 1). For all 3 timepoints, source and the interaction
with cage significantly explained most of the observed community
variation (combined R\textsuperscript{2} = 0.90, 0.99, 0.88,
respectively; \emph{P} = 0.0001; Fig. 3 and Table S6). We also compared
baseline communities across the 2 experiments, and found experiment and
cage significantly explained the observed variation only for the Schloss
and Young lab mouse colonies (Fig. S2 and Table S7), suggesting vendor
communities were relatively stable between the 2 experiments. Thus,
mouse colony source was the factor that explained the most variation
observed in the bacterial communities. Importantly with the exception of
the 2 University colonies, the community of each source clustered apart
from one another suggesting each community had a unique response to
clindamycin treatment and \emph{C. difficile} challenge.

After finding differences at the community level, we next identified the
taxa that varied across sources of mice over the initial days of the
experiment. We examined bacterial relative abundances at the operational
taxonomic unit (OTU) and family levels, expecting the number of
differences to be reduced at the family level due to the nature of
bacterial taxonomy (45). Focusing on the baseline communities first,
there were 268 OTUs and only 20 families (Table S8-9) with relative
abundances that varied across colony sources. Clindamycin treatment
reduced the number of taxa with relative abundances that varied across
sources to 18 OTUs and 10 families (Table S8-9). After \emph{C.
difficile} challenge, there were 44 OTUs and 18 families (Table S8-9)
with significantly different relative abundances across sources, as the
communities started to recover from antibiotic treatment. In spite of
the experimental perturbations that occurred during these 3 timepoints,
there were 12 OTUs (Fig 4A-C) and 8 families with relative abundances
that consistently varied across colony sources (Fig. 4D-F). Importantly,
some of the taxa that consistently varied across sources also shifted
with clindamycin treatment. For example, \emph{Proteus} increased after
clindamycin treatment, but only in Taconic mice. \emph{Enterococcus} was
primarily found only in mice purchased from commercial vendors and also
increased after clindamycin treatment. In summary, mouse bacterial
communities significantly varied according to colony source throughout
the course of the experiment and a consistent subset of bacterial taxa
remained different across sources regardless of clindamycin and \emph{C.
difficile} challenge.

\textbf{Clindamycin treatment alters a subset of taxa that were found in
all colony sources.} Although there were bacteria that consistently
varied across colony sources, we also wanted to identify the bacteria
that shifted after clindamycin treatment, regardless of colony source.
By analyzing all mice that had sequence data from fecal samples
collected at baseline and after clindamycin treatment, we identified 153
OTUs and 18 families that were altered after clindamycin treatment (Fig.
5 and Table S10-11). Interestingly, when we compared the list of
significant clindamycin impacted bacteria with the bacteria that
consistently varied across groups over the initial 3 timepoints of our
experiment, we found 3 OTUs (\emph{Lachnospiraceae} (OTU 130),
\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) and 3
families (\emph{Porphyromonadaceae}, \emph{Enterococcaceae},
\emph{Lachnospiraceae}) overlapped (Fig. 4, Fig. 5C-D). These findings
demonstrate that clindamycin has a consistent impact on the fecal
bacterial communities of mice from all colony sources and only a subset
of the taxa also varied across colony sources.

\textbf{Source-specific and clindamycin impacted bacteria distinguish
\emph{C. difficile} colonization status in mice.} After identifying taxa
that varied by colony source, changed after clindamycin treatment, or
both, we next wanted to determine which taxa were influencing the
variation in \emph{C. difficile} colonization status at day 7 (Fig. 1D).
We trained L2-regularized logistic regression models with input
bacterial community data from the baseline, post-clindamycin, and
post-infection timepoints of the experiment to predict \emph{C.
difficile} colonization status on day 7 (Fig. S3). All models were
better at predicting \emph{C. difficile} colonization status on day 7
than random chance (all \emph{P} \(\le\) 5e-15; Table S12), however the
models trained with OTU level data generally performed better than those
trained with family level data with the expception of the models based
on the post-infection (day 1) communities (Fig. 6). Interestingly, the
model based on the post-clindamycin (day 0) community OTU data performed
significantly better than all other models with an AUROC of 0.75
(\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for pairwise comparisons;
Table S13). Thus, we were able to use community bacterial relative
abundance data alone to differentiate mice that were still colonized
with \emph{C. difficile} on day 7 from those mice that had cleared the
\emph{C. difficile}. Interestingly, the model built with OTU relative
abundance data post-clindamycin treatment had the best performance,
suggesting how the bacterial community responds to clindamycin treatment
has the greatest influence on subsequent \emph{C. difficile}
colonization dynamics.

Next, to examine the bacteria that were driving each model's
performance, we pulled out the top 20 taxa that had the highest absolute
feature weights in each of the 6 models (Table S14-15). The top 20 taxa
from each model were compared to the list of taxa that varied across
colony source (Fig. 4 and Table S8-9) at the same timepoint and the taxa
that were altered by clindamycin treatment (Fig. 5 and Table S10-11). We
found a subset of OTUs and families that were important to the model and
overlapped with bacteria that varied by either source, clindamycin
treatment, or both (Fig. S4, S5, S6). Combining the overall results for
the 3 OTU models identified 14 OTUs associated with source, 21 OTUs
associated with clindamycin treatment, and 6 OTUs associated with both
(Fig. 7A). Combining the overall results for the 3 family models
identified 18 families associated with source, 14 families associated
with clindamycin treatment and 8 families associated with both (Fig.
7B). Several OTUs (\emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)}) and families
(\emph{Bacteroidaceae, Deferribacteraceae, Enterococcaceae,
Lachnospiraceae, Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
Verrucomicrobiaceae}) appeared across at least 2 models, so we examined
how the relative abundances of these key taxa varied over the course of
the experiment (Fig. 8 and Fig. S7). Throughout the experiment, there
was at least 1 timepoint where relative abundances of these taxa
significantly varied across sources (Table S16-17). Interestingly, there
were no taxa that emerged as consistently enriched or depleted in
Jackson, Taconic, and Envigo mice that were colonized for longer with
\emph{C. difficile} 630, suggesting multiple bacteria influence the time
needed to clear the infection. Together, these results suggest the
initial bacterial communities and their responses to clindamycin have a
large influence on the time needed to clear \emph{C. difficile}.

@@

By examining the \emph{C. difficile} colonization dynamics within mice
from 6 different colony sources after perturbing the microbiota with
clindamycin treatment, we were able to identify bacterial taxa that were
unique to sources throughout the experiment as well as taxa that were
universally impacted by clindamycin. We built L2 logistic regression
models with baseline, post-clindamycin treatment, and post-infection
fecal community data that successfully predicted \emph{C. difficile}
colonization status 7 days after infection better than random chance. We
identified \emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)},
\emph{Bacteroidaceae, Deferribacteraceae, Enterococcaceae,
Lachnospiraceae, Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
Verrucomicrobiaceae} (Fig. 8, Fig. S7) as candidate bacteria within
these communities that were influencing variation in \emph{C. difficile}
colonization dynamics since these bacteria were all important in the
logistic regression models and varied by colony source, were impacted by
clindamycin treatment, or both. Overall, our results demonstrate
clindamycin is sufficient to render mice from multiple sources
susceptible to CDI and only a subset of the interindividual microbiota
variation across mice from different sources was associated with the
time needed to clear \emph{C. difficile}.

Other groups have taken similar approaches by using mice from multiple
colony sources to identify bacteria that either promote colonization
resistance or increase susceptibility to enteric infections (22, 23,
26--30). For example, in the context of \emph{Salmonella} infections,
\emph{Enterobacteriaceae} and segmented filamentous bacteria have
emerged as protective (22, 27). A previous study with \emph{C.
difficile} identified an endogenous protective \emph{C. difficile}
strain LEM1 that bloomed after antibiotic treatment in mice from Jackson
or Charles River Laboratories, but not Taconic that protected mice
against the more toxigenic \emph{C. difficile} VPI10463 (26). Given that
we ordered mice from the same vendors, we checked all mice for
endogenous \emph{C. difficile} by plating stool samples that were
collected after clindamycin treatment. However, we did not identify any
endogenous \emph{C. difficile} strains prior to challenge, suggesting
there were no endogenous protective strains in the mice we received and
other bacterial taxa were mediating the variation in \emph{C. difficile}
colonization across sources. Although all mice were susceptible to
\emph{C. difficile} colonization, by following colonization over time we
found Jackson, Taconic, and Envigo mice remained colonized for longer.
We identified a subset of bacteria that were important in predicting
whether a mouse was still colonized with \emph{C. difficile} status 7
days post-infection. These results suggest a subset of the bacterial
community is responsible for determining the length of time needed to
clear \emph{C. difficile} colonization.

>>>>

The microbiota plays a key role in determining susceptibility to
\emph{Clostridioides difficile} infections (CDIs). However, much of the
mechanistic work examining CDIs in mouse models use a single university
colony or vendor. We treated mice from 6 different colony sources (2
University of Michigan colonies and 4 vendors) with a single clindamycin
dose, followed by \emph{C. difficile} challenge 1 day later and measured
\emph{C. difficile} colonization levels through 9 days post-infection.
The microbiota was profiled via 16S rRNA gene sequencing analysis to
examine variation across colony sources and alterations due to
clindamycin treatment and \emph{C. difficile} challenge. While all
sources of mice were colonized 1-day post-infection, variation in
\emph{C. difficile} colonization levels emerged from days 3-7
post-infection with 3 sources colonized with \emph{C. difficile} for
slightly longer and at higher levels. We identified bacterial taxa with
different relative abundances across colony sources throughout the
experiment, as well as taxa that were consistently impacted by
clindamycin treatment in all sources of mice. We created logistic
regression models that successfully classified mice based on whether
they cleared \emph{C. difficile} by 7 days post-infection using
baseline, post-clindamycin, and post-infection community composition
data. After examining the taxa that were most important to the
classification models, we identified a subset of key taxa that varied
across colony sources (\emph{Bacteroides}, \emph{Deferribacteraceae}),
were altered by clindamycin (\emph{Porphyromonadaceae},
\emph{Ruminococcaceae}), or both (\emph{Enterobacteriaceae},
\emph{Enterococcus}, \emph{Bifidobacteriaceae},
\emph{Coriobacteriaceae}, \emph{Lachnospiraceae}, and
\emph{Verrucomicrobiaceae}). These results suggest the response of the
initial gut microbiota to clindamycin treatment influences \emph{C.
difficile} 630 colonization dynamics.

@@

Mouse models of CDIs have been a great tool for understanding \emph{C.
difficile} pathogenesis (13). The number of CDI mouse model studies has
grown substantially since Chen et al. published their C57BL/6 model in
2008, which disrupted the gut microbiota with antibiotics to enable
\emph{C. difficile} colonization and symptoms such as diarrhea and
weight loss (14). CDI mouse models have been used to examine
translationally relevant questions regarding \emph{C. difficile},
including the role of the microbiota and efficacy of potential
therapeutics for treating CDIs (15). However, microbiome variation
between lab mice is much less than the variation observed between humans
(16, 17). Additionally, studying the contribution of the microbiota to a
particular disease phenotype in one set of lab mice after the same
perturbation could yield a number of findings of which only a fraction
may be driving the phenotype.

@@

\textbf{Clindamycin treatment renders all mice susceptible to \emph{C.
difficile} 630 colonization regardless of colony source.} To test how
the microbiotas of mice from different colony sources impact
colonization dynamics after clindamycin exposure, we utilized C57BL/6
mice from 6 different sources: two colonies from the University of
Michigan (the Young and Schloss lab colonies), the Jackson Laboratory,
Charles River Laboratories, Taconic Biosciences, and Envigo (which was
formerly Harlan). These 4 vendors were chosen because they represent
commonly used vendors for CDI studies in mice (26, 34--40). After a
13-day acclimation period for the mice ordered from vendors, all mice
were treated with 10 mg/kg clindamycin via intraperitoneal injection and
one day later challenged with 10\textsuperscript{3} \emph{C. difficile}
630 spores (Fig. 1A). Clindamycin was chosen because we have previously
demonstrated mice are rendered susceptible, but consistently cleared the
CDI within 9 days (21, 41), clindamycin is frequently implicated with
human CDIs (42), and is also part of the antibiotic treatment for the
frequently cited 2008 CDI mouse model (14). The day after infection,
\emph{C. difficile} was detectable in all mice at a similar level
(median CFU range: 2.2e+07-1.3e+08; \emph{P}\textsubscript{FDR} = 0.15),
indicating clindamycin rendered all mice susceptible regardless of
colony source (Fig. 1B). Interestingly, variation in \emph{C. difficile}
CFU levels across sources of mice emerged from days 3-7 post-infection
(all \emph{P}\textsubscript{FDR} \(\le\) 0.019; Fig. 1B and Table S1),
suggesting mouse colony source is associated with \emph{C. difficile}
clearance. We conducted two experiments approximately 3 months apart and
while the colonization dynamics were similar across most sources of
mice, there was some variation between the 2 experiments, particularly
for the Schloss and Envigo mice (Fig. S1A-B, Fig. 1D). Although \emph{C.
difficile} 630 causes mild symptoms in mice comparied to other \emph{C.
difficile} strains (43), we also saw that weight change significantly
varied across sources of mice with the most weight lost two days
post-infection (Fig. 1C, E and Table S2). Interestingly, mice ordered
from Jackson, Taconic, and Envigo tended to lose more weight (although
there was variation between experiments with Schloss and Envigo mice),
have higher \emph{C. difficile} CFU levels and take longer to clear the
infection compared to the other sources of mice, which was particularly
evident 7 days post-infection (Fig. 1B-E), when 57.50 of the mice were
still colonized with \emph{C. difficile} (Fig. S1C). By 9 days
post-infection the majority of the mice from all sources had cleared
\emph{C. difficile} (Fig. S1C) with the exception of 1 Taconic mouse
from the first experiment and 2 Envigo mice from the second experiment.
Importantly, there was also one Jackson and one Envigo mouse that died
between 1- and 3-days post-infection during the second experiment. Thus,
clindamycin rendered all mice susceptible to \emph{C. difficile} 630
colonization, regardless of colony source, but variation across sources
emerged with 3 out of 6 sources taking longer to clear \emph{C.
difficile}.

\textbf{Bacterial communities consistently vary across mouse colony
sources despite antibiotic and infection perturbations.} Given the well
known variation in mouse microbiomes across vendors and university
colonies (25), we hypothesized that the variation in \emph{C. difficile}
clearance could be explained by microbiota variation across the 6
sources. We used 16S rRNA gene sequencing to characterize the fecal
bacterial communities from the mice over the course of the experiment.
Since antibiotics and other risk factors of CDIs are associated with
decreased microbiota diversity (44), we first examined alpha diversity
measures across the 6 sources of mice. Examining the bacterial
communities at baseline, prior to clindamycin treatment there was a
significant difference in the number of observed OTUs
(\emph{P}\textsubscript{FDR} = 0.03), but not Shannon diversity index
(\emph{P}\textsubscript{FDR} = 0.052) across sources of mice (Fig. 2A-B
and Table S3). As expected, clindamycin treatment decreased richness and
Shannon diversity across all sources of mice, and communities started to
recover 1 day post-infection (Fig. 2C-D). Interestingly, significant
differences in diversity metrics across sources
(\emph{P}\textsubscript{FDR} \textless{} 0.05) emerged after both
clindamycin and \emph{C. difficile} infection, with Charles River mice
having higher richness and Shannon Diversity than most of the other
groups (Fig 2C-F and Table S4). While Charles River mice had more
diverse microbiotas, Young and Schloss lab mice were also able to clear
\emph{C. difficile} faster, suggesting microbiota diversity alone does
not explain the observed variation in \emph{C. difficile} colonization
across vendors.

Next, we compared the bacterial communities from the 6 colonies over the
course of the experiment using principal coordinate analysis (PCoA) of
the Theta YC distances. Permutational multivariate analysis of variance
(PERMANOVA) analysis revealed colony source was the major factor
explaining the observed variation across fecal communities
(R\textsuperscript{2} = 0.35, \emph{P} = 0.0001) followed by
interactions between cage and day of the experiment (Movie S1 and Table
S5). Since, the majority of the perturbations happened over the initial
days of the experiment, we decided to focus on the bacterial communities
at baseline (day -1), after clindamycin treatment (day 0), and
post-infection (day 1). For all 3 timepoints, source and the interaction
with cage significantly explained most of the observed community
variation (combined R\textsuperscript{2} = 0.90, 0.99, 0.88,
respectively; \emph{P} = 0.0001; Fig. 3 and Table S6). We also compared
baseline communities across the 2 experiments, and found experiment and
cage significantly explained the observed variation only for the Schloss
and Young lab mouse colonies (Fig. S2 and Table S7), although most of
the vendors also clustered by experiment, suggesting there was some
community variation between the 2 experiments within each vendor. Thus,
mouse colony source was the factor that explained the most variation
observed in the bacterial communities. Importantly with the exception of
the 2 University colonies, the community of each source clustered apart
from one another suggesting each community had a unique response to
clindamycin treatment and \emph{C. difficile} challenge.

Since there was some variation in microbiota communities between
experiments at baseline, we next looked at how similar the communities
were within the same source and between sources in response to
clindamycin treatment and \emph{C. difficile} challenge (Fig. 4) . The
baseline communities varied most between experiments for Schloss, Young,
and Envigo mice and variation between sources of mice was high (Fig.
4A). Clindamycin treatment reduced variation between experiments within
Schloss, Young and Jackson mice and some of the variation between
sources diminished (Fig. 4B). Post-infection, the community variation
started to increase within sources of mice and variation between sources
of mice started to return (Fig. 4C). By using mice from multiple sources
we were able to increase the number of microbiota communities we tested
with the clindamycin \emph{C. difficile} colonization mouse model.

After finding differences at the community level, we next identified the
taxa that varied across sources of mice over the initial days of the
experiment. We examined bacterial relative abundances at the operational
taxonomic unit (OTU) and family levels, expecting the number of
differences to be reduced at the family level due to the nature of
bacterial taxonomy (45). Focusing on the baseline communities first,
there were 268 OTUs and 20 families (Table S8-9) with relative
abundances that varied across colony sources. Clindamycin treatment
reduced the number of taxa with relative abundances that varied across
sources to 18 OTUs and 10 families (Table S8-9). After \emph{C.
difficile} challenge, there were 44 OTUs and 18 families (Table S8-9)
with significantly different relative abundances across sources, as the
communities started to recover from antibiotic treatment. In spite of
the experimental perturbations that occurred during these 3 timepoints,
there were 12 OTUs (Fig 5A-C) and 8 families with relative abundances
that consistently varied across colony sources (Fig. 5D-F). Importantly,
some of the taxa that consistently varied across sources also shifted
with clindamycin treatment. For example, \emph{Proteus} increased after
clindamycin treatment, but only in Taconic mice. \emph{Enterococcus} was
primarily found only in mice purchased from commercial vendors and also
increased after clindamycin treatment. In summary, mouse bacterial
communities significantly varied according to colony source throughout
the course of the experiment and a consistent subset of bacterial taxa
remained different across sources regardless of clindamycin and \emph{C.
difficile} challenge.

\textbf{Clindamycin treatment alters a subset of taxa that were found in
all colony sources.} Although there were bacteria that consistently
varied across colony sources, we also wanted to identify the bacteria
that shifted after clindamycin treatment, regardless of colony source.
By analyzing all mice that had sequence data from fecal samples
collected at baseline and after clindamycin treatment, we identified 153
OTUs and 18 families that were altered after clindamycin treatment (Fig.
6 and Table S10-11). Interestingly, when we compared the list of
significant clindamycin impacted bacteria with the bacteria that
consistently varied across groups over the initial 3 timepoints of our
experiment, we found 3 OTUs (\emph{Lachnospiraceae} (OTU 130),
\emph{Lactobacillus} (OTU 6), \emph{Enterococcus} (OTU 23)) and 3
families (\emph{Porphyromonadaceae}, \emph{Enterococcaceae},
\emph{Lachnospiraceae}) overlapped (Fig. 5, Fig. 6C-D). These findings
demonstrate that clindamycin has a consistent impact on the fecal
bacterial communities of mice from all colony sources and only a subset
of the taxa also varied across colony sources.

\textbf{Source-specific and clindamycin impacted bacteria distinguish
\emph{C. difficile} colonization status in mice.} After identifying taxa
that varied by colony source, changed after clindamycin treatment, or
both, we next wanted to determine which taxa were influencing the
variation in \emph{C. difficile} colonization at day 7 (Fig. 1D, Fig.
S1C). We trained L2-regularized logistic regression models with input
bacterial community data from the baseline, post-clindamycin, and
post-infection timepoints of the experiment to predict \emph{C.
difficile} colonization status on day 7 (Fig. S3A-B). All models were
better at predicting \emph{C. difficile} colonization status on day 7
than random chance (all \emph{P} \(\le\) 5e-15; Table S12), however the
models trained with OTU level data generally performed better than those
trained with family level data with the expception of the models based
on the post-infection (day 1) communities (Fig. S3C-D). Interestingly,
the model based on the post-clindamycin (day 0) community OTU data
performed significantly better than all other models with an AUROC of
0.75 (\emph{P}\textsubscript{FDR} \(\le\) 3.9e-10 for pairwise
comparisons; Table S13). Thus, we were able to use community bacterial
relative abundance data alone to differentiate mice that had cleared
\emph{C. difficile} before day 7 from the mice still colonized with
\emph{C. difficile}. Interestingly, the model built with OTU relative
abundance data post-clindamycin treatment had the best performance,
suggesting how the bacterial community responds to clindamycin treatment
has the greatest influence on subsequent \emph{C. difficile}
colonization dynamics.

Next, to examine the bacteria that were driving each model's
performance, we pulled out the top 20 taxa that had the highest absolute
feature weights in each of the 6 models (Table S14-15). First, we looked
at OTUs from the model with the best performance that was based on the
post-clindamycin treatment bacterial community data. While most of the
20 OTUs had low relative abundances on day 0, \emph{Enterobacteriace},
\emph{Bacteroides} and \emph{Proteus} had high relative abundances in at
least one source of mice and significantly varied across sources (Fig.
7A). Next, the top 20 taxa from each model were compared to the list of
taxa that varied across colony source (Fig. 5 and Table S8-9) at the
same timepoint and the taxa that were altered by clindamycin treatment
(Fig. 6 and Table S10-11). We found a subset of OTUs and families that
were important to the model and overlapped with bacteria that varied by
either source, clindamycin treatment, or both (Fig. S4, S5A-C).
Combining the overall results for the 3 OTU models identified 14 OTUs
associated with source, 21 OTUs associated with clindamycin treatment,
and 6 OTUs associated with both (Fig. 7B). Combining the overall results
for the 3 family models identified 18 families associated with source,
14 families associated with clindamycin treatment and 8 families
associated with both (Fig. S5D). Several OTUs (\emph{Bacteroides (OTU
2), Enterococcus (OTU 23), Enterobacteriaceae (OTU 1),
Porphyromonadaceae (OTU 7)}) and families (\emph{Bacteroidaceae,
Deferribacteraceae, Enterococcaceae, Lachnospiraceae,
Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
Verrucomicrobiaceae}) appeared across at least 2 models, so we examined
how the relative abundances of these key taxa varied over the course of
the experiment (Fig. 8 and Fig. S6). Throughout the experiment, there
was at least 1 timepoint where relative abundances of these taxa
significantly varied across sources (Table S16-17). Interestingly, there
were no taxa that emerged as consistently enriched or depleted in mice
that were colonized past 7 days post-infection with \emph{C. difficile}
630, suggesting multiple bacteria influence the time needed to clear the
infection. Together, these results suggest the initial bacterial
communities and their responses to clindamycin have a large influence on
the time needed to clear \emph{C. difficile}.

@@

By examining the \emph{C. difficile} colonization dynamics within mice
from 6 different colony sources after perturbing the microbiota with
clindamycin treatment, we were able to identify bacterial taxa that were
unique to sources throughout the experiment as well as taxa that were
universally impacted by clindamycin. We built L2 logistic regression
models with baseline, post-clindamycin treatment, and post-infection
fecal community data that successfully predicted \emph{C. difficile}
colonization status 7 days after infection better than random chance. We
identified \emph{Bacteroides (OTU 2), Enterococcus (OTU 23),
Enterobacteriaceae (OTU 1), Porphyromonadaceae (OTU 7)},
\emph{Bacteroidaceae, Deferribacteraceae, Enterococcaceae,
Lachnospiraceae, Bifidobacteriaceae, Coriobacteriaceae, Ruminococcaceae,
Verrucomicrobiaceae} (Fig. 8, Fig. S6) as candidate bacteria within
these communities that were influencing variation in \emph{C. difficile}
colonization dynamics since these bacteria were all important in the
logistic regression models and varied by colony source, were impacted by
clindamycin treatment, or both. Overall, our results demonstrate
clindamycin is sufficient to render mice from multiple sources
susceptible to CDI and only a subset of the interindividual microbiota
variation across mice from different sources was associated with the
time needed to clear \emph{C. difficile}.

Other groups have taken similar approaches by using mice from multiple
colony sources to identify bacteria that either promote colonization
resistance or increase susceptibility to enteric infections (22, 23,
26--30). For example, in the context of \emph{Salmonella} infections,
\emph{Enterobacteriaceae} and segmented filamentous bacteria have
emerged as protective (22, 27). A previous study with \emph{C.
difficile} identified an endogenous protective \emph{C. difficile}
strain LEM1 that bloomed after antibiotic treatment in mice from Jackson
or Charles River Laboratories, but not Taconic that protected mice
against the more toxigenic \emph{C. difficile} VPI10463 (26). Given that
we ordered mice from the same vendors, we checked all mice for
endogenous \emph{C. difficile} by plating stool samples that were
collected after clindamycin treatment. However, we did not identify any
endogenous \emph{C. difficile} strains prior to challenge, suggesting
there were no endogenous protective strains in the mice we received and
other bacterial taxa mediated the variation in \emph{C. difficile}
colonization across sources. Although all mice were susceptible to
\emph{C. difficile} colonization, by following colonization over time we
found Jackson, Taconic, and Envigo mice remained colonized beyond 7 days
post-infection. We identified a subset of bacteria that were important
in predicting whether a mouse was still colonized with \emph{C.
difficile} 7 days post-infection. These results suggest a subset of the
bacterial community is responsible for determining the length of time
needed to clear \emph{C. difficile} colonization.

In the past variation between different CDI mouse model studies have
been attributed to intestinal microbiome differences in mice across
different institutional environments. For example, groups using the same
clindamycin treatment and C57BL/6 mice had different \emph{C. difficile}
outcomes, one having sustained colonization (32), while the other had
transient (18). Baseline differences in the microbiota composition have
been hypothesized to partially explain the differences in colonization
outcomes and overall susceptibility to \emph{C. difficile} after
treatment with the same antibiotic (13, 31). We have shown that mice
from 6 different sources were all susceptible to \emph{C. difficile}
630, suggesting the microbiota influences \emph{C. difficile} clearance
more than susceptibility. Fortunately, the bacterial perturbations
induced by clindamycin treatment have been well characterized and our
findings agree with previous CDI mouse model work demonstrating
\emph{Enterococcus} and \emph{Enterobacteriaceae} were associated with
\emph{C. difficile} susceptibility and \emph{Porpyhromonadaceae},
\emph{Lachnospiraceae}, \emph{Ruminococcaceae}, and \emph{Turicibacter}
were associated with resistance (19, 21, 32, 33, 41, 46--48). While we
have demonstrated that susceptibility is uniform across vendors after
clindamycin treatment, there could be different outcomes for either
susceptibility or clearance in the case of other antibiotic treatments.
The \emph{C. difficile} strain used could also be contributing to the
variation in \emph{C. difficile} outcomes seen across different groups
(47). We found the time needed to naturally clear \emph{C. difficile}
varied across sources of mice implying that at least in the context of
the same perturbation, microbiota differences seemed to influence
infection outcome more than susceptibility. More importantly, we were
able to narrow down from all the variation observed across colony
sources to a subset of bacterial taxa that were also important for
predicting \emph{C. difficile} colonization status 7 days
post-infection. Since all but 3 mice eventually cleared \emph{C.
difficile} 630 by 9 days post-infection and the model built with the
post-clindamycin OTU relative abundance data had the best performance,
our results suggest clindamycin treatment had a large role in
determining \emph{C. difficile} susceptibility and clearance in the
mice.
